Rheology of Cubic Blue Phases

### 2013-07-03 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2039 40876441-3496-0410-a613-a2fca35b2ae2

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
}
Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. 
\rev{
Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states.
}
For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, 
and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ v_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{op-eom}
\end{equation}
The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity advected by the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
and is explicitly given by~\cite{Beris:1994}
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q W}),
\label{sw}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution for $\theta$, the 
so-called Leslie angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that respects the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{molfield}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

@@

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma} l$. 
\rev{
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
\rev{
Depending on the configuration in steady shear flow five different regimes 
can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

@@

\com{Still need to check:} 
\rev{
Surprisingly, if the shear is switched off in
the BPI-2 regime at some point during an oscillation, 
the flow-induced configuration usually
reverts to a perfect quiescent BPI -- in stark contrast with
what happens in BPI-1 for lower shear rates or BPI-3 for higher shear rates. 
}

@@

\rev{
At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
periodically recurring disclination network in regime BPI-2 and the formation
of another amorphous network. A sequence of snapshots of this is shown in Fig. \ref{bp1-3-disc}. 
At first the disclination lines take the form of a regular, staggered array of helices 
with the helical axis lying along the flow direction.
However, this mode of flow proves equally unstable as the regular formations dissolve into 
an irregular state with helicoidal disclination lines of varying 
appearance. The morphology is different from the amorphous network of
regime BPI-1 (cf. Fig. \ref{disc-bp1-1}) where the disclination lines are 
well separated but do not have a clear orientation with respect to the flow. 
}

@@

\rev{
In all our runs we used Lees-Edwards boundary conditions \cite{Wagner:2002}.
This is the sheared equivalence of periodic boundary conditions, for both the 
velocity and the order parameter field. 
On crossing a Lees-Edwards boundary a Galilean transformation is applied
with a velocity increment which is fixed in time (and limited in size by the
low Mach number constraint of LB). An appropriate transformation
of LB distributions which propagate across a boundary
must be applied at each time step, and appropriate
adjustment to the tensor of velocity gradients $W_{\alpha\beta}$ 
is required to compute cross-plane gradients of the velocity field 
used in the update to ${\mathbf Q}$. In both cases interpolation of the
relevant quantities is required to cope with the relative displacement
of neighbouring lattice sites separated by a sliding plane (the
displacement may be a fraction of a lattice unit at any given time
step) \cite{Henrich:2012a}. The use of multiple sliding planes equally spaced in a single
system allows the overall shear rate to be maintained indefinitely as the system
becomes larger in the velocity gradient direction (in contrast with the use of solid
walls to impose shear).
While extremely useful to simulate bulk flow, we
note that these boundary conditions impose macroscopic distortion
of the network. They do not allow for a free-standing network as
in flow-gradient direction there can be no slip in the network deformation 
or strain at any point.
This is effectively equivalent to infinitely distant
walls along the $xy-$plane where the liquid crystal is anchored. 
Effectively, this disallows permeation flows, which on the other
hand require a non-zero velocity field and, simultaneously,
much lower (ideally zero) perturbations of the blue phase network.
Consequently our results are effectively valid for sheared bulk 
BPs which are anchored at the walls. In practice, permeation flows are 
only restricted to very low flow rates, and are unstable for intermediate 
and fast flows~\cite{Dupuis:2005}, 
where the response of the network depends less on
the details of the boundary conditions used at the wall.
With free boundary conditions and for very low flow velocity, 
permeative flows might instead allow for
some slip of the BPI network, as in cholesterics sheared
along their axis~\cite{Marenduzzo:2006a, Marenduzzo:2006b}.
}

>>>>

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
}
Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- 
\rev{
very slow flow causes a transition of the initially 
ordered structure with oscillatory response into an amorphous network 
with an apparent yield stress. 
Larger shear rates lead to another amorphous state with different structure of 
the defect network.
}
For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, 
and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ v_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{op-eom}
\end{equation}
The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity advected by the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
and is explicitly given by~\cite{Beris:1994}
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q \mathbf W}),
\label{sw}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution for $\theta$, the 
so-called Leslie angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that respects the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{molfield}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

@@

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma}\, l$. 
\rev{
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{
\rev{
Flow curve $\bar{\Pi}_{xy}(\gd)$: 
Depending on the configuration in steady shear flow five different regimes 
can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

@@

\com{This is currently under investigation:} 
\rev{
Surprisingly, if the shear is switched off in
the BPI-2 regime at some point during an oscillation, 
the flow-induced configuration can revert to a perfect quiescent BPI -- in stark contrast with
what happens in BPI-1 for lower shear rates or BPI-3 for higher shear rates.
We also see the network being stuck in metastable states with 
helicoidal disclinations similar to the bottom left picture in Fig.~\ref{bp1-1-disc} 
and the top left image in Fig.~\ref{bp1-3-disc}.
}

@@

\rev{
At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
periodically recurring disclination network in regime BPI-2 and the formation
of another amorphous network. A sequence of snapshots of this is shown in Fig. \ref{bp1-3-disc}. 
At first the disclination lines take the form of a regular, staggered array of helices 
with the helical axis lying along the flow direction.
However, this mode of flow proves equally unstable as the regular formations dissolve into 
an irregular state with helicoidal disclination lines of varying 
appearance. The morphology is different from the amorphous network of
regime BPI-1 (cf. Fig.~\ref{bp1-1-disc}) where the disclination lines are 
well separated but do not have a clear orientation with respect to the flow. 
}

@@

\rev{
In all our runs we used Lees-Edwards boundary conditions (LEBCs)~\cite{Wagner:2002}.
This is the sheared equivalence of periodic boundary conditions, for both the 
velocity and the order parameter field. 
On crossing a Lees-Edwards boundary a Galilean transformation is applied
with a velocity increment which is fixed in time (and limited in size by the
low Mach number constraint of LB). An appropriate transformation
of LB distributions which propagate across a boundary
must be applied at each time step, and appropriate
adjustment to the tensor of velocity gradients $W_{\alpha\beta}$ 
is required to compute cross-plane gradients of the velocity field 
used in the update to ${\mathbf Q}$. In both cases interpolation of the
relevant quantities is required to cope with the relative displacement
of neighbouring lattice sites separated by a sliding plane (the
displacement may be a fraction of a lattice unit at any given time
step) \cite{Henrich:2012a}. It is worth mentioning that apart from 
the Galilean transformation LEBCs do not impose any further constraint.
This means that order parameter field is free to follow its own kinetic pathway
at the Lees-Edwards plane. 
The use of multiple sliding planes equally spaced in a single
system allows the overall shear rate to be maintained indefinitely as the system
becomes larger in the velocity gradient direction (in contrast with the use of solid
walls to impose shear).
While extremely useful to simulate bulk flow, we
note that these boundary conditions impose macroscopic distortion
of the network. They do not allow for a free-standing network as
in flow-gradient direction there can be no slip in the network deformation 
or strain at any point.
The boundary condition is equivalent to infinitely distant
walls along the $xy-$plane where the liquid crystal is anchored,
not unlike the situation found in a rheometer with no-slip
boundary conditions at the walls.
Effectively, this disallows permeation flows, which on the other
hand require a non-zero velocity field and simultaneously
much lower (ideally zero) perturbations of the blue phase network.
In practice, permeation flows are 
only restricted to very low flow rates, and are unstable for intermediate 
and fast flows~\cite{Dupuis:2005}, 
where the response of the network depends less on
the details of the boundary conditions used at the wall.
With free boundary conditions that do not impose a macroscopic distortion 
and for very low flow velocity, 
permeative flows might instead allow for
some slip of the BPI network, as in cholesterics sheared
along their axis~\cite{Marenduzzo:2006a, Marenduzzo:2006b}.
}
